---
title: "LungMAP Project Lung Development"
author: "Harsh Bhotika; Geremy Clair"
date: "May 20, 2020"
output: html_document
html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme : darkly
    code_folding: hide
---

```{r knit_options, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=6, echo = TRUE, results = TRUE, warning = FALSE, message=FALSE) 
options(repos=structure(c(CRAN="https://mirror.ibcp.fr/pub/CRAN/")))
```

# Loading the data(Gene list) and the required Libraries
The data contained in the file "Gene List Full.csv" containing a list of 258 genes for this search were loaded into the R environment.
```{r load_data}
genes <- read.csv(file.choose(), header = TRUE)
View(genes)
attach(genes)
data.class(genes)
summary(genes)
```

Loading the necessary Libraries (Note: If you updated R recently, then the latest version of these have to be re-installed.).
```{r libraries}
library(RISmed)
library(magrittr)
library(dplyr)
library(ggplot2)
library(tidytext)
library(wordcloud)
```

# Load functions to fetch Publication data 
Function to fetch publication data from the PubMed database. Change parameters "retmax", "mindate" and "maxdate" as per requirements.
```{r fetch_data}
fetch_gene_data<- function(gene=x, key_word= "alveolarization", exact_match=T){ 
  if(missing(exact_match)){exact_match<-TRUE}
  key_word<-tolower(key_word)
  # if you get some object error, either reduce retmax or try removing it.
  res1 <- EUtilsSummary(paste0(gene,' , "' , key_word,'"'), 
                        type = "esearch", 
                        db = "pubmed",
                        datetype = "pdat",
                        retmax = 1000,
                        mindate = 1970, 
                        maxdate = 2020)
  fetch <- EUtilsGet(res1, type = "efetch", db = "pubmed")
  
  abstracts <- data.frame(title = fetch@ArticleTitle,
                          abstract = fetch@AbstractText, 
                          journal = fetch@Title,
                          pmid = fetch@PMID, 
                          year = fetch@YearPubmed)
  
  #if(exact_match==T){
   # position <- agrep(key_word, paste(abstracts$title, abstracts$abstract, sep = ". "))
    #abstracts <- abstracts[position[],]
    #rownames(abstracts)=NULL}
  
  abstracts <- abstracts %>% mutate(abstract = as.character(abstract))
  abstracts %>%
    head()
  return(abstracts)}
```

Function to run the search and compile the publication data returned in every search iteration.
```{r compile_data}
query_gene_keyword_Pubmed<-function(gene_list=x, key_word=y, exact_match=TRUE, waiting_time=0){
  if(missing(exact_match)){exact_match<-TRUE}
  if(missing(waiting_time)){waiting_time<-0}
  gene_list<-as.character(t(gene_list)) # trick to make sure data frame converts to char
  
  final_data <- NULL
  
  for (i in 1:length(gene_list)){
    x<- gene_list[i]
    abstract_data <- fetch_gene_data(x, key_word=key_word, exact_match=T)
    if(nrow(abstract_data)>0){
    abstract_data<- cbind(gene_name=x,abstract_data)
    final_data <- rbind(final_data, abstract_data)}
    Sys.sleep(waiting_time)
  }
  
  return(final_data)
}
```

# Initiate the search and export the results
Pass the required parameters in the function call below to initiate the search as per requirements i.e. change the "key_word" which is currenty set to "Lung development". (Note: The search takes time as the data lookup is huge).
```{r initiate_search}
export_data<-query_gene_keyword_Pubmed(genes,key_word = "Lung development", waiting_time = 5)
```

View and export the compiled data. Data will be saved in the default folder.
```{r }
View(export_data)
write.csv(export_data, file= "PubMed_data.csv")
```

#Additional Analyses
Plot a graph of the results by year. Change axis labels depending on the search paramaters above
```{r plot_graph}
export_data %>%
  group_by(year) %>%
  count() %>%
  filter(year > 1970) %>%
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  labs(title = "Pubmed Articles over the Years (1970 to 2020)", hjust = 1,
       y = "Articles")
```

Create a word cloud of the words in the abstracts of the final results. Change parameters as per requirements.
```{r wordcloud}
cloud <- export_data %>%
  unnest_tokens(word, abstract) %>%
  anti_join(stop_words) %>%
  count(word, sort = TRUE)

cloud %>%
  with(wordcloud(word, n, min.freq = 2, max.words = 100, colors = brewer.pal(8, "Dark2")), scale = c(8,.3), per.rot = .4)
```
Every PubMed Id has been manually curated to ensure a proper attribution of the keywords.