---
title: 'Trend_Analysis'
author: Lisa Bramer, Geremy Clair
date: April 19, 2021
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme : cerulean
    code_folding: hide
---

Set the knitR options
```{r knit_options, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=6, echo = FALSE, results = TRUE, warning = FALSE, message=FALSE) 
options(repos=structure(c(CRAN="https://cloud.r-project.org/")))
```

Load the necessary libraries
```{r libraries}
library(pmartR)
library(ggplot2)
library(reshape2)
library(tidyr)
library(purrr)
library(dplyr)
library(ggpubr)
library(DT)
library(datadr)
```

Set the working directory
```{r directory}
setwd("/Volumes/Lungmap/19 - Biopsie paper/Code/01_Trend_analysis/")
```

Read the peptide data
```{r pep_data}
pepdata = read.csv("./01_source files/TMT_Peptides_with_column_header.csv")
```

Create PmartR e_data 
```{r e_data}
edata = pepdata[,c(2,5:ncol(pepdata))]
```

Parse column names to match meta data file #
```{r parse_colnames}
names(edata)[-1] = unlist(lapply(strsplit(names(edata)[-1], "_"), function(x) x[1]))
```

Pull out e_meta 
```{r e_meta}
emeta = pepdata[,1:4]
```

Read in f_data 
```{r f_data}
fdata = read.csv("./01_source files/TMT_metadata.csv")
```

Create some extra columns related to age 
```{r age_cols}
fdata$Age_Group = c("Days", "Days", "Days", "Days", "Months", "Months", "Months", "Years", "Years", "Years", "Years")
fdata$Age_Days = c(1,1,7,14,30.44, 4*30.44, 7*30.44, 365.242, 2*365.242, 4*365.242, 8*365.242)
fdata$Ln_Age_Days_plus1 = log(fdata$Age_Days+1)
```

# save data as pepData object 
```{r create_pepData}
mydata = as.pepData(e_data = edata, f_data = fdata, e_meta = emeta, edata_cname = "Peptide", fdata_cname = "Donor", emeta_cname = "Protein")
```

replace 0's with NA's
```{r zeros_to_NA}
mydata = edata_replace(omicsData = mydata, x = 0, y = NA)
```

generate a qqplot prior normalization
```{r qqplot_pre_norm}
m<-mydata$e_data[,2:ncol(mydata$e_data)]
meltedm<-melt(m)
meltedm<-meltedm[!is.na(meltedm$value),]
ggqqplot(meltedm$value)+ggtitle("qqplot_prior_transformation")
```


Evaluate normality prior normalization using a Wilk shapiro test on 5000 proteins drawned randomly for each sample (R Shapiro test is limited to 5000 entities)
```{r normality_test_prior}
# draw random rows in the data
set.seed(7)
r<-sample.int(nrow(mydata$e_data),5000)
# extract data
m<-mydata$e_data[r,2:ncol(mydata$e_data)]

#run Wilk_shapiro normality test
SW_test<-lapply(m,shapiro.test)

#extract W and pvalue
W<-numeric()
p<-numeric()
s<-paste0(colnames(m),"_prior_log_norm")
for(i in 1:length(SW_test)){
  W[i]<-SW_test[[i]]$statistic
  p[i]<-SW_test[[i]]$p.value
}

pre_norm_SW<-data.frame(sample=s,W=W,p.value=p)
print("the prenormalization table Wilk-shapiro test results are displayed below")
datatable(pre_norm_SW)

```

replace zeros by NAs
```{r NA_replacement}
mydata = edata_replace(omicsData = mydata, x = 0, y = NA)
```

log2 transform the data, generate the qqplot and a distribution histogram for each sample
```{r log2_transform}
mydata_log2 = edata_transform(mydata, "log2")
m<-mydata_log2$e_data[,2:ncol(mydata_log2$e_data)]
meltedm<-melt(m)
meltedm<-meltedm[!is.na(meltedm$value),]
ggqqplot(meltedm$value)+ggtitle("qqplot_prior_transformation")
ggplot(meltedm,aes(value))+geom_histogram(bins=100)+ggtitle("global data frequency histrogram")
ggplot(meltedm,aes(value))+geom_histogram(bins=100)+ggtitle("data frequency histogram (per sample)")+facet_wrap(~variable)
ggplot(meltedm,aes(x=variable, y=value))+geom_boxplot(aes(fill=variable))+ggtitle("boxplot (per sample)")
```

Evaluate normality after log_transform using a Wilk shapiro test on 5000 proteins drawned randomly for each sample (R Shapiro test is limited to 5000 entities)
```{r normality_test_after}
# extract data (using the same draw as prior)
m<-mydata_log2$e_data[r,2:ncol(mydata_log2$e_data)]

#run Wilk_shapiro normality test
SW_test<-lapply(m,shapiro.test)

#extract W and pvalue
W<-numeric()
p<-numeric()
s<-paste0(colnames(m),"_post_norm")
for(i in 1:length(SW_test)){
  W[i]<-SW_test[[i]]$statistic
  p[i]<-SW_test[[i]]$p.value
}

post_norm_SW<-data.frame(sample=s,W=W,p.value=p)

#display results and plot the prior and after normalization WS test W values
SW<-rbind(pre_norm_SW, post_norm_SW)
print("the pre-normalization and post-normalization table Wilk-shapiro test results prior and after norm")
datatable(SW)
SW$sample<-factor(SW$sample, levels =SW$sample)

ggplot(SW,aes(x=W,y=sample))+geom_bar(stat="identity")+ggtitle("W from Wilk-shapiro test")

```

remove contaminants
```{r contaminant_rm}
# find proteins that are contiaminants #
prots_rmv = pepdata$Protein[grep("Contaminant", pepdata$Protein)]
# contaminant filter #
contam_filt = custom_filter(e_meta_remove = as.character(prots_rmv), omicsData = mydata_log2)
# apply the contaminant filter #
mydata1 = applyFilt(contam_filt, mydata_log2)
```

remove misidentifications
```{r missidentifications_rm}
# find proteins that are misidentifications #
mis_ids = grep("XXX", mydata1$e_meta$Protein, value = T)
# create filter for misidentified proteins #
misid_filt = custom_filter(mydata1, e_meta_remove = mis_ids)
# apply misidentification filter #
mydata2 = applyFilt(misid_filt, mydata1)
```

create filter to remove proteins with less than two peptides
```{r}
# create filter to remove proteins with too few peptides
myfilt = proteomics_filter(mydata2)
# remove proteins with only one peptide observed
mydata3 = applyFilt(myfilt, mydata2, min_num_peps = 2)
```

create filter to remove proteins with too few peptides
```{r low_pep_filt}
# create filter to remove proteins with too few peptides
myfilt = proteomics_filter(mydata2)
# remove proteins with only one peptide observed
mydata3 = applyFilt(myfilt, mydata2, min_num_peps = 2)
```

create filter to remove peptides observed only in one sample
```{r low_count_filt}
# create filter to remove peptides based on non-missing values
myfilt2 = molecule_filter(mydata3)
# remove peptides observed in only one sample
mydata4 = applyFilt(myfilt2, mydata3)
```

perform a median nomalization
```{r median_norm}
# normalize the data #
norm_data = normalize_global(mydata4, subset_fn = "all", norm_fn = "median", apply_norm = T, backtransform = T)
m<-norm_data$e_data[,2:ncol(norm_data$e_data)]
meltedm<-melt(m)
meltedm<-meltedm[!is.na(meltedm$value),]
ggplot(meltedm,aes(value))+geom_histogram(bins=100)+ggtitle("data frequency histogram (per sample)")+facet_wrap(~variable)
ggplot(meltedm,aes(x=variable, y=value))+geom_boxplot(aes(fill=variable))+ggtitle("boxplot (per sample)")

```

rollup the peptide information to the protein level
```{r rrollup}
prodata = protein_quant(norm_data, method = "rrollup", use_parallel = F)
```

run group designation
```{r}
prodata_grouped = group_designation(prodata, main_effects = "Age_Group")
```

run sppca
```{r}
pca_grouped = dim_reduction(prodata_grouped)
full_pca = pcaMethods::pca(t(as.matrix(prodata_grouped$e_data[,-1])), method = "ppca", scale = "vector")
```

prepare data for statistics
```{r}
## prepare data for statistics
melt_edata = melt(prodata$e_data, id.var = "Protein")
names(melt_edata) = c("Protein", "Donor", "value")
peps_per_pro = norm_data$e_meta %>% group_by(Protein) %>% summarise(N = n())
melt_all = merge(x = melt_edata, y = prodata$f_data, by = "Donor")
melt_all2 = merge(x = melt_all, y = peps_per_pro, by = "Protein", all.x = T, all.y = F)
# divide the data by protein 
byProt = melt_all2 %>% group_by(Protein)
# filter out proteins without enough data to fit a model
byProt_filt = byProt %>% dplyr::filter(sum(!is.na(value)) > 4)
```


Create a function to calculate the stats
```{r stat_fn}
      stat_fn = function(x){
        # log transform age #
        x$log_age = log(x$Age_Days + 1)
        
        # fit linear and quadratic models #  
        lin_mod = lm(value ~ log_age, data = x)
        quad_mod = lm(value ~ poly(log_age,2), data = x)
        
        if(summary(lin_mod)$coeff[2,4] < 0.1){
          trend_0.1 = "Linear"
        }else{trend_0.1 = NA}
        if(summary(quad_mod)$coeff[3,4] < 0.1){
          trend_0.1 = "Quadratic"
        }
        
        if(summary(lin_mod)$coeff[2,4] < 0.05){
          trend_0.05 = "Linear"
        }else{trend_0.05 = NA}
        if(summary(quad_mod)$coeff[3,4] < 0.05){
          trend_0.05 = "Quadratic"
        }
        
        
        lm_lin_p = summary(lin_mod)$coeff[2,4]
        lm_lin_e = summary(lin_mod)$coeff[2,1]
        qm_quad_p = summary(quad_mod)$coeff[3,4]
        qm_quad_e = summary(quad_mod)$coeff[3,1]
        
        # fit loess model #
        loess_mod = loess(value ~ log(Age_Days), data = x, span = 1)
        loess_fitted = predict(loess_mod, newdata = data.frame(Age_Days = 1:2922))
        loess_slps = diff(loess_fitted)
        loess_dir = ifelse(loess_slps > 0, 1, -1)
        loess_dir_changes = sum(loess_dir[-1] != loess_dir[-length(loess_dir)], na.rm = T)
        if(!is.na(trend_0.1)){
          if(trend_0.1 == "Quadratic"){loess_dir_changes = 1}
          if(trend_0.1 == "Linear"){loess_dir_changes = 0}
        }
        
        data.frame(
          N_peptides = x$N[1],
          N_samples = sum(!is.na(x$value)), 
          N_direction_changes = loess_dir_changes,
          LM_linear_pvalue = lm_lin_p, 
          LM_slope_estimate = lm_lin_e,
          QM_quadratic_pvalue = qm_quad_p,
          QM_quadratic_estimate = qm_quad_e
        )
        
      }
```


perform the stats
```{r perform_stats}
# process the data
byProt_stat_res = byProt_filt %>% nest() %>% mutate(stat_res = map(data, stat_fn))
# pull out statistics results
stat_results = byProt_stat_res %>% dplyr::select(Protein, stat_res) %>% unnest(stat_res)
```

Determine the best fitted trend
```{r determine_best_fitted_trend}
stat_results$trend = "None"
for(i in 1:nrow(stat_results)){
  if(stat_results$LM_linear_pvalue[i]<0.05|stat_results$QM_quadratic_pvalue[i]<0.05){
    if(stat_results$LM_linear_pvalue[i]<stat_results$QM_quadratic_pvalue[i]){
      if(stat_results$LM_slope_estimate[i]>0){stat_results$trend[i]<-"linear_increasing"}else{stat_results$trend[i]<-"linear_decreasing"}
    }else{
      if(stat_results$QM_quadratic_estimate[i]>0){stat_results$trend[i]<-"quadratic_convex"}else{stat_results$trend[i]<-"quadratic_concave"}
    }
  }
}
par(mar=c(10,10,5,5))
barplot(table(stat_results$trend)[-3],xlab="trend",las=2)
```


